Effect of initial volume fraction on the collapse of granular columns in fluid

### 2018-08-26 paper/paper.tex
:pencil: Summary
:pencil: Conclusions
:pencil: Comparison between loose v dense

\section{UNDERWATER GRANULAR FLOWS}
In this study, a 2D poly-disperse system ($d_{max}/d_{min} = 1.8$) of circular discs in fluid was used to understand the behaviour of granular flows on inclined planes (see~\Cref{fig:setup}). The soil column was modelled using 1000 discs of density \SI{2650}{\kg\per\cubic\meter} and a contact friction angle of \SI{26}{\degree}. The collapse of the column was simulated inside a fluid with a density of \SI{1000}{\kg\per\cubic\meter}  and a kinematic viscosity of \SI{1e-6}{\square\meter\per\second}. The choice of a 2D geometry has the advantage of cheaper computational effort than a 3D case, making it feasible to simulate very large systems. A granular column of aspect ratio `a' of 0.8 was used. A hydrodynamic radius r = 0.9R was adopted during the LBM computations. Dry analyses were also performed to study the effect of hydrodynamic forces on the run-out distance.

@@

\citeNP{Rondon2011} observed that the loose packings flow rapidly on a time
scale proportional to the initial height and results in longer run-out distance 
in comparison to the dense packing. Hydroplaning occurs above a critical 
Froude's number of 0.4. The Froude's number is inversely related to the 
thickness of the flow and its density. Hence, for the same thickness of flow, a 
loose granular column will experience more hydroplaning than a dense granular 
flow. This effect might result in longer run-out behaviour in fluid than the 
dry condition for the same initial aspect ratio. The initial packing density 
and the permeability of a 2D granular column, with an aspect ratio of 0.8, are 
varied to understand their influence on the run-out behaviour. The run-out 
behaviour of the dense case (83\% packing density), discussed in the previous 
section, is compared with a loose granular column (79\% packing fraction). The 
permeability is varied by changing the hydrodynamic radius from 0.7 \textit{R} 
to 0.95 
R.

@@

\citeNP{Rondon2011} also observed that the collapse of a granular column in a 
viscous fluid is mainly controlled by the initial volume fraction and not by 
the aspect ratio of the column. The role of the initial volume fraction 
observed explains the pore pressure feedback mechanism proposed 
by~\citeNP{Schaeffer2008,Iverson2000} in the context of landslides. The 
compaction or dilation of grains can cause additional stress in the grains 
which can stabilise or destabilise the soil. The flow is thus controlled by the 
coupling between the dilatancy of the granular layer and the development of 
pore pressure in the fluid phase~\cite{Pailha2008}. The dense column needs to 
dilate in order to flow. When it starts to fall, liquid is then sucked into the 
column, which is then stabilised by the additional viscous 
drag~\cite{Rondon2011,Topin2012}. By opposition the loose column when 
it starts flowing expands and ejects liquid, leading to a partial fluidisation 
of the material.

>>>>

\section{UNDERWATER GRANULAR FLOWS}
In this study, a 2D poly-disperse system ($d_{max}/d_{min} = 1.8$) of circular discs in fluid was used to understand the behaviour of granular flows on inclined planes (see~\Cref{fig:setup}). The soil column was modelled using 1000 discs of density \SI{2650}{\kg\per\cubic\meter} and a contact friction angle of \SI{26}{\degree}. The collapse of the column was simulated inside a fluid with a density of \SI{1000}{\kg\per\cubic\meter}  and a kinematic viscosity of \SI{1e-6}{\square\meter\per\second}. The choice of a 2D geometry has the advantage of cheaper computational effort than a 3D case, making it feasible to simulate very large systems. A granular column of aspect ratio `a' of 0.8 was used. Dry analyses were also performed to study the effect of hydrodynamic forces on the run-out distance.

@@

\citeNP{Rondon2011} observed that the loose packings flow rapidly on a time
scale proportional to the initial height and results in longer run-out distance 
in comparison to the dense packing. Hydroplaning occurs above a critical 
Froude's number of 0.4. The Froude's number is inversely related to the 
thickness of the flow and its density. Hence, for the same thickness of flow, a 
loose granular column will experience more hydroplaning than a dense granular 
flow. This effect might result in longer run-out behaviour in fluid than the 
dry condition for the same initial aspect ratio. The initial packing density 
and the permeability of a 2D granular column, with an aspect ratio of 0.8, are 
varied to understand their influence on the run-out behaviour. The run-out 
behaviour of the dense case (83\% packing density), is compared with a loose
granular column (79\% packing fraction). The permeability is varied by changing
the hydrodynamic radius from 0.7 \textit{R} to 0.95 \textit{R}.

The normalised run-out for different hydrodynamic radii for a granular column 
with an initial aspect ratio of 0.8 for dense initial packing are presented 
in~\cref{fig:Runout_a08_dense}. The run-out increases with decrease in the 
permeability, which is equivalent to an increase in the hydrodynamic radius. 
An increase in the hydrodynamic radius from 0.7 to 0.95 \textit{R} increases 
the normalised run-out by 25\%. However, even under a very low permeability 
condition (\textit{r} = 0.95 \textit{R}), the run-out observed in fluid is 
shorter than the dry and the buoyant conditions.

\begin{figure}[htpb]
\centering
\includegraphics[width=0.9\linewidth]{figs/Runout_a08_dense}
\caption{Effect of permeability on the evolution of run-out for a column 
collapse in fluid (a = 0.8).}
\label{fig:Runout_a08_dense}
\end{figure}

At a high permeability (\textit{r} = 0.7 \textit{R}), the evolution of run-out 
at the initial stage is quicker, which means that the negative pore-pressure 
that is developed during the shearing along the shear-failure surface is 
dissipated faster. Even though the negative pore-pressure is dissipated, due to 
the development of negative pore-pressure the evolution of run-out in fluid is 
slower than its dry counterpart. The rate of pore-pressure dissipation 
decreases with decrease in the permeability. This can be observed by a flatter 
slope in the run-out evolution with decrease in the 
permeability.~\Cref{fig:PWP_ini_dense} shows the distribution of pore-pressure 
in high and low permeability granular media along the horizontal direction at a 
height of $10 \times d$ from the base. In LBM, the pore-pressure in fluid is a 
function of fluid density distribution functions. At time $t = \tau_c$, the 
highly permeable (\textit{r} = 0.7 \textit{R}) granular column 
shows smaller negative pore-pressure in comparison to large negative 
pore-pressures observed in the shearing zone of a low permeable column 
(\textit{r} = 0.9 R). This shows that not only does it take longer for the 
pore-pressure to dissipate with a decrease in permeability, but also results in 
almost twice the negative pore-pressure than what is observed in the high 
permeable case (\cref{fig:r095_PWP_ini_dense,fig:r095_PWPress_ini_dense}). A 
high value of positive pore-pressure is observed at the face of the column 
in the low permeable condition. This indicates that the low permeable column 
fails as a continuous block undergoing a shear failure, which generates a very 
large negative pore-pressure along the failure surface. However, in the case of 
the high permeable column the failure is more localised with multiple negative 
pore-pressure spikes (\cref{fig:r07_PWPress_ini_dense}).

\begin{figure}
\centering
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_ini_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r07_PWP_ini_dense}
\end{subfigure} \\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_ini_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r095_PWP_ini_dense}
\end{subfigure}
\end{figure}
\begin{figure}
\ContinuedFloat
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_ini_press_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:r07_PWPress_ini_dense}
\end{subfigure} \\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_ini_press_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:r095_PWPress_ini_dense}
\end{subfigure}

\caption{Effect of permeability on the excess pore water pressure distribution 
for a granular column collapse in fluid (a = 0.8 \& dense packing) at $t = 
\tau_c$ along the horizontal direction at a height of 10d from the base.}
\label{fig:PWP_ini_dense}
\end{figure}

Although the low permeability granular columns require a longer duration for 
the run-out to evolve, the final run-out distance is found to be much longer 
than in the high permeable condition.~\Cref{fig:PE_a08_dense} shows that the 
potential energy mobilised during the flow of a low permeability column is 20\% 
smaller than the collapse of a high permeability granular column. Normalised 
kinetic energy evolution with time (\cref{fig:a08_dense_energy}) shows that the 
low permeability column has a wider peak kinetic energy distribution in 
comparison to a sharp peak observed in the high permeability condition. This 
indicates the influence of lubrication, i.e., hydroplaning of the granular flow 
in low permeability conditions. The evolution of the horizontal kinetic energy 
with time reveals that the peak kinetic energy is sustained longer as the 
permeability of the granular material decreases (\cref{fig:KExy_a08_dense}). 
Although the peak kinetic energy is smaller in the low permeability case, the 
hydroplaning of the flowing granular mass results in a longer run-out 
distance.~\Cref{fig:r095_PWP_flow_dense,fig:r095_PWPress_flow_dense} shows the 
distribution of pore-pressure for a dense granular column collapse in fluid 
along the bottom plane. A high positive pore-pressure is observed 
at the base of the granular flow at the flow front in low permeability 
condition indicating the occurrence of hydroplaning. The positive pore-pressure 
at the flow front decreases the effective stress thus the creating lubrication 
effect. The evolution of local packing density with time shows that the packing 
density decreases with decrease in permeability 
(\cref{fig:Packing_Density_a08_dense}). This drop 
in the value of packing density between $t = 2\tau_c$ and $t=3\tau_c$ 
corroborates with the duration of hydroplaning during which a large amount of 
water in entrained at the flow front.

\begin{figure}
	\centering
    \includegraphics[width=0.7\linewidth]{figs/PE_a08_dense}
    \caption{Effect of permeability on the evolution of the potential energy 
    with time for a granular column collapse in fluid (a = 0.8).}
    \label{fig:PE_a08_dense}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.8\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/KE_a08_dense}
    \caption{Evolution of the total kinetic energy.}
    \label{fig:KE_a08_dense}
\end{subfigure}
}\\
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.95\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/KExy_a08_dense}
    \caption{Evolution of horizontal and vertical kinetic energies.}
    \label{fig:KExy_a08_dense}
\end{subfigure}
}
\caption{Effect of permeability on the evolution of kinetic energies with time 
for a granular column collapse in fluid (a = 0.8).}
\label{fig:a08_dense_energy}
\end{figure}

\begin{figure}
\centering
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_flow_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure at the 
    bottom of the granular flow.}
    \label{fig:r07_PWP_flow_dense}
\end{subfigure}
\\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_flow_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure at the 
        bottom of the granular flow.}
    \label{fig:r095_PWP_flow_dense}
\end{subfigure}
\end{figure}
\begin{figure}
\ContinuedFloat
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r07_PWP_flow_press_dense}
    \caption{High permeability (\textit{r} = 0.7 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:r07_PWPress_flow_dense}
\end{subfigure}
\\
\begin{subfigure}[t]{0.975\linewidth}
	\centering
    \includegraphics[width=\linewidth]{figs/a08/r095_PWP_flow_press_dense}
    \caption{Low permeability (\textit{r} = 0.95 \textit{R}) - Pressure contour 
    (Pa).}
    \label{fig:r095_PWPress_flow_dense}
\end{subfigure}
\caption{Effect of permeability on the excess pore water pressure distribution 
along the bottom plane for a granular column collapse in fluid (a = 0.8 \& 
dense packing) at $t = 2\tau_c$.}
\label{fig:PWP_flow_dense}
\end{figure}

\begin{figure}
\centering
\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.95\linewidth}
	\centering
    \includegraphics[width=0.8\linewidth]{figs/Packing_Density_a08_dense}
    \caption{Evolution of the packing density.}
    \label{fig:Packing_Density_a08_dense}
\end{subfigure}
}\\

\makebox[\linewidth][c]{
\begin{subfigure}[t]{0.95\linewidth}
	\centering
    \includegraphics[width=0.8\linewidth]{figs/Fr_a08_dense}
    \caption{Evolution of the Froude's number.}
    \label{fig:Fr_a08_dense}
\end{subfigure}
}
    \caption{Effect of permeability on the evolution of packing density and 
    Froude's number for a granular column collapse in fluid (a = 0.8 \& dense 
    initial packing).}
    \label{fig:Packing_Density_Fr_a08_dense}
\end{figure}

The high permeability granular column shows lower water entrainment 
(\cref{fig:Packing_Density_a08_dense}), which indicates that for highly 
permeable flows the drag force acting on the soil grains predominates over the 
lubrication effect on the run-out behaviour.~\Cref{fig:Fr_a08_dense} shows the 
evolution of Froude's number with time for different permeability conditions. 
In dense granular columns, the Froude's number increases with decrease in 
permeability, however the Froude's number is below a critical value of 0.4 even 
for a low permeable column. Hence, no hydroplaning is observed. In both the low 
and high permeable granular flows, the granular material consolidates at the 
final stage of the flow (\cref{fig:Packing_Density_a08_dense}). This can be 
observed by the increase in the packing density at the final stage due to 
settlement of grains and expulsion of entrained water. The final deposit 
profile for both low and high permeability conditions are shown 
in~\cref{fig:a08_dense_snapshots}. The high permeability collapse shows a more 
parabolic (convex) deposit profile in contrast to the more concave profile 
observed in the low permeability condition. In high permeable condition, once 
the granular mass is mobilised, the drag forces slows down the rate of flow, 
which allows the grains to densify resulting in a more parabolic profile. This 
is in contrast to the hydroplaning observed in the low permeability condition, 
where the granular mass thickness is smaller than the dense condition. The 
observation of hydroplaning in the low permeable condition may be due to the 
difference in the distribution of granular mass at the flow front.

@@

\section{Summary}

Two-dimensional LB-DEM simulations were performed to understand the behaviour of
submarine granular flows. Unlike dry granular collapse, the run-out behaviour in
fluid is dictated by the initial volume fraction.
\citeNP{Rondon2011} also observed that the collapse of a granular column in a 
viscous fluid is mainly controlled by the initial volume fraction and not by 
the aspect ratio of the column. The role of the initial volume fraction 
observed explains the pore pressure feedback mechanism proposed 
by~\citeNP{Schaeffer2008,Iverson2000} in the context of landslides. The 
compaction or dilation of grains can cause additional stress in the grains 
which can stabilise or destabilise the soil. The flow is thus controlled by the 
coupling between the dilatancy of the granular layer and the development of 
pore pressure in the fluid phase~\cite{Pailha2008}. The dense column needs to 
dilate in order to flow. When it starts to fall, liquid is then sucked into the 
column, which is then stabilised by the additional viscous 
drag~\cite{Rondon2011,Topin2012}. By opposition the loose column when 
it starts flowing expands and ejects liquid, leading to a partial fluidisation 
of the material.
